---
title: "Design-Based Latent Class Measurement in Complex Surveys"
subtitle: "Carrying replicate variance from the measurement model through to domain estimates"
author: "Kevin Linares"
date: today
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    fig-width: 7
    fig-height: 4.5
    fig-align: center
    df-print: kable
    geometry: margin=1in
    fig-crop: false
bibliography: bibliography.bib
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup}
#| include: false
# The nine packages library(tidyverse) would attach, named rather than pulled in
# through the meta-package, so that a function running here runs unchanged in
# the application that implements this specification.
pacman::p_load(lavaan, matrixStats, clue, survey, furrr, kableExtra,
               viridis, here, glue, sjlabelled, haven, knitr, janitor,
               dplyr, purrr, tibble, tidyr, stringr, ggplot2, readr,
               forcats, lubridate)

source("source_code.R")

# The method configuration is sourced into its own environment rather than
#   into this one, so the object names it defines -- cfg, items, cats -- cannot
#   collide with anything here and every chunk below says which arm it means.
#
# One subtlety, stated because it is invisible otherwise. The configuration
#   pulls in the shared data reader with source(), which evaluates in the global
#   environment rather than in the environment sys.source() was given, so the
#   objects that reader defines -- recode_audit among them -- arrive as globals
#   rather than in the returned list. Worth an assertion rather than a
#   discovery.
load_arm <- function(path) {
  e <- new.env(parent = globalenv())
  sys.source(path, envir = e, keep.source = FALSE)
  as.list(e)
}
arm_lca <- load_arm("survey_lca_config.R")

stopifnot(exists("recode_audit"), exists("demos"))

# The dimension of each model is an analyst decision recorded in configuration,
#   not a quantity this paper selects. @sec-enumeration presents the evidence;
#   this is the number that was entered after reading it. A paper reports a
#   completed analysis, so an unset dimension is an error here rather than a
#   section that waits.
if (is.null(arm_lca$cfg$K_force))
  stop("Set cfg$K_force in survey_lca_config.R before rendering. The ",
       "enumeration in @sec-enumeration is the evidence that decision is ",
       "taken from.", call. = FALSE)

K_star <- arm_lca$cfg$K_force
```

# Introduction

Survey batteries are read one item at a time far more often than they are read
together. Where they are read together, the latent variable model that does it
is usually fitted as though the respondents had been drawn one at a time and
independently, which almost none of them were. This paper sets out a workflow
in which the sample design enters every stage of a latent variable analysis:
the measurement model is estimated under the design weights, its parameters
carry design-based standard errors, and the domain estimates that follow
inherit their variance from the same replicate design rather than from a fresh
derivation.

The gap this addresses is narrow but consequential. Design-weighted point
estimation for latent class and factor models is available in several packages
and in Mplus. What is not available in one place is uncertainty that respects
stratification and clustering, carried without interruption from the
measurement model through to the corrections applied to the domain estimates
that depend on it. Analysts therefore report loadings and conditional
probabilities with model-based standard errors that assume simple random
sampling, and then compare groups on quantities derived from them. Where the
design effect on those parameters is materially above one, the resulting
intervals do not have their nominal coverage, and differences are called
significant that the data cannot support.

Two models are treated here rather than one, because the choice between them is
part of the methodology rather than prior to it. A latent class model asks
which of a small number of kinds of respondent a person resembles; a factor
model asks where on a continuum they sit. Both are measurement models for the
same battery, both rest on local independence, and which one applies is a
property of the data. Presenting them together makes the shared machinery
visible: one replicate design, one variance routine, and one domain estimation
step serve both, and only the measurement model and the scoring step differ.

Placed in the total survey error framework [@groves2009; @groves2010], the
workflow addresses one component and one only. Sampling error, including the
adjustment-error cost that weight variation imposes, is carried faithfully from
the measurement model to the domain estimate. Measurement error in the latent
variable --- classification error in one arm, score error in the other --- is
corrected rather than assumed away. Everything upstream is untouched:
coverage, unit nonresponse, and the quality of the weights that adjust for it
are taken as given, and processing error in the recodes is made visible by an
audit rather than removed by a method. Being specific about which component is
being handled is what keeps a design-based standard error from being read as a
warrant for the whole estimate.

The contribution is not new estimation theory. The weighted pseudo-likelihood
for latent class analysis is due to @patterson2002 and @vermunt2007; the
treatment of ordinal indicators through polychoric correlations and a
weighted-least-squares estimator is standard [@rosseel2012]; replicate variance
for complex designs is @wolter2007 and @rust1996. What is offered is their
composition, the design-based diagnostics that composition makes possible, and
an explicit account of the points at which the workflow can mislead --
including the points where an assertion had been standing in for a
computation.

# Two models, and the question that chooses between them {#sec-choosing}

Both models take a battery of categorical answers and posit something
underneath that the individual items measure imperfectly. They differ in what
that something is assumed to be.

Let $\mathbf{y}_i = (y_{i1},\dots,y_{iJ})$ be respondent $i$'s answers to $J$
items, item $j$ having $C_j$ ordered categories. The latent class model with
$K$ classes, prevalences $\pi_k$ and conditional response probabilities
$\rho_{jck} = P(y_{ij} = c \mid X_i = k)$ states

$$P(\mathbf{y}_i) = \sum_{k=1}^{K} \pi_k \prod_{j=1}^{J} \rho_{j\,y_{ij}\,k},
\qquad \sum_{k}\pi_k = 1, \qquad \sum_{c}\rho_{jck} = 1 .$$ {#eq-lca}

The factor model instead posits a continuous $\eta_i$ and, for each item, an
unobserved continuous response $x^*_{ij}$ cut into categories by thresholds
$\tau_{jc}$,

$$x^*_{ij} = \lambda_j \eta_i + \varepsilon_{ij},
\qquad y_{ij} = c \iff \tau_{j,c-1} < x^*_{ij} \le \tau_{jc} .$$ {#eq-cfa}

@eq-lca is the classical latent structure model [@lazarsfeld1968;
@goodman1974]. The product over items in it and the independence of the
$\varepsilon_{ij}$ in @eq-cfa are the same assumption under two names: **local independence**,
that conditional on the latent variable the items carry no further relationship
to one another. It is what allows either model to be summarised so compactly,
and it is the first thing to fail when a battery contains two questions that
are really asking the same thing. @sec-diagnostics probes it in both arms.

The assumptions that differ are these. The class model treats the response
categories as unordered, so it can recover a group that avoids the middle of a
scale or clusters on one particular answer, and it does not require that more
of anything means a higher answer. The factor model requires exactly that
monotonicity, and in exchange estimates far fewer quantities: a five-class
model on thirteen seven-category items carries several hundred free parameters,
a one-factor model on the same items fewer than a hundred. An item that
violates monotonicity can separate classes strongly while contributing almost
nothing to a factor, and the reverse also occurs; where the two arms disagree
about an item, the disagreement describes the battery rather than indicating
that one model is wrong.

The practical question is whether respondents differ in *how much* or in
*which*. We make it answerable rather than rhetorical with a level-to-pattern
ratio. For a fitted class model, scale each item's expected response within
class $k$ to the unit interval,

$$m_{jk} = \frac{\sum_{c} c\,\rho_{jck} - 1}{C_j - 1},$$ {#eq-scaled}

so that a binary and a seven-point item contribute the same possible spread;
without that rescaling the ratio is an artefact of the response formats. Write
$\bar m_k$ for the mean of $m_{jk}$ over items. The ratio is

$$R = \frac{\mathrm{sd}_{jk}\!\left(m_{jk} - \bar m_k\right)}
          {\mathrm{sd}_{k}\!\left(\bar m_k\right)} .$$ {#eq-ratio}

Well below one says the classes differ mainly in how high they answer overall,
which is a continuum a factor model would describe with far fewer parameters.
Near or above one says they reorder the items, which is structure no single
factor can hold. It is evidence, not a test, and it is read alongside what the
analyst knows about the questions.

This paper, and the application that implements it, fit the class model only.
The ratio is retained because the question it answers still has to be asked: a
battery whose classes differ mainly in level is a continuum, and fitting
classes to it produces an ordered staircase that looks like a finding and is an
artefact of the model. Where the ratio comes back well below one the right
response is a factor model, which is a different specification and not
something this workflow provides.

# Data and preparation {#sec-data}

The demonstration uses the 2023 AmericasBarometer round for Ecuador, conducted
by the LAPOP Lab at Vanderbilt University [@lapop2023]. Fieldwork was carried
out face to face in Spanish by IPSOS between February and April 2023, with a
multi-stage probability sample stratified by the three major regions, and
released base weights. Two batteries are analysed, chosen to make the contrast
of @sec-choosing concrete: a set of items on the material circumstances of the
household and the respondent's own situation, of mixed response format across
unrelated domains, and a set of institutional trust and system support items on
a single seven-point ladder.

Preparation follows three rules, and each exists because its absence is silent
rather than loud.

**Nonresponse codes are recoded to missing with the source file read so that
they are visible.** SPSS declares user-missing values in the file; reading with
`user_na = TRUE` keeps them as values rather than letting them arrive as `NA`,
so the analyst decides whether a "don't know" is a category or a missing value
rather than inheriting the reader's default. A code that is nonresponse on one
item and a substantive answer on another is the hazard here, and it is one the
analyst has to be shown before the decision is taken.

**Items are recoded to consecutive integers on every row of the file**, with
the levels taken from the rows that will actually be fitted. The estimation
frame and the prediction frame then share one coding by construction, and a
value seen only outside the estimation set becomes missing and drops out of
that respondent's product rather than silently indexing a different category.

**Demographic recodes are exhaustive and halt on an unmatched label.** This is
the one place in the workflow where a mistake produces no symptom: nothing
fails, no statistic looks wrong, and every later table is quietly about the
wrong people. Every observed source label is given a target group explicitly,
and the audit of what became what is a table to be read once per dataset.

```{r recode-audit}
# recode_audit is a global rather than a member of either arm: see the note in
#   the setup chunk on where the shared reader's objects land.
recode_audit |>
  lca_table(col.names = c("Variable", "Source label", "Recoded to", "n"),
            widths = c("6em", "21em", "9em", "3em"),
            caption = "Recode audit: what every source label became. Read once
                       per dataset; a misgrouping here is invisible downstream.")
```

```{r batteries}
tibble(Item = arm_lca$items,
       Question = str_trunc(arm_lca$dictionary$question, 60),
       Categories = arm_lca$cats) |>
  lca_table(align = "llr", widths = fit_widths(3, first = 1.4), font_size = 7,
            caption = "The battery. Mixed response formats across unrelated
                       domains, which is the case a class model is built for.")
```

# Estimation under a complex design {#sec-estimation}

## The design-weighted pseudo-likelihood

With base weights $w_i$, both arms maximise a pseudo-likelihood rather than a
likelihood. For the class model,

$$\ell_w(\boldsymbol\theta) = \sum_{i=1}^{n} w_i
\log\!\Big(\sum_{k=1}^{K}\pi_k \prod_{j=1}^{J}\rho_{j\,y_{ij}\,k}\Big),$$ {#eq-pseudo}

the standard device under complex sampling [@patterson2002; @vermunt2007;
@asparouhov2005; @skinner1989]. @eq-pseudo is the sample-based estimate of the
census log-likelihood that would be computed if the whole population were
observed. Its maximiser is consistent for the census parameter under standard
conditions, but it is not the maximiser of a probability model for the observed
data. That distinction is not pedantic: it is the reason the design must supply
the variance rather than the curvature of @eq-pseudo, and it is why the
information matrix is not used for standard errors anywhere in this workflow.

The EM algorithm [@dempster1977] alternates an expectation step,

$$p_{ik} = \frac{\pi_k \prod_j \rho_{j\,y_{ij}\,k}}
{\sum_{l}\pi_l \prod_j \rho_{j\,y_{ij}\,l}},$$ {#eq-estep}

and a maximisation step in which the weights enter every sufficient statistic,

$$\hat\pi_k = \frac{\sum_i w_i p_{ik}}{\sum_i w_i},
\qquad
\hat\rho_{jck} = \frac{\sum_i w_i\, p_{ik}\, \mathbf 1(y_{ij}=c)}
{\sum_i w_i\, p_{ik}\, \mathbf 1(y_{ij}\ \text{observed})} .$$ {#eq-mstep}

Three implementation points are load-bearing. Restricting the denominator in
@eq-mstep to respondents who answered item $j$ is what keeps
$\sum_c\hat\rho_{jck} = 1$ under item nonresponse; with complete cases the
restriction is vacuous, and where it is not, omitting it produces conditional
distributions that do not sum to one. The expectation step accumulates
$\log\rho$ across items and applies the log-sum-exp identity before
exponentiating, because with many items the raw product underflows to zero and
@eq-estep becomes $0/0$. And a missing answer contributes zero on the log
scale, which is exactly equivalent to dropping it from the product in
@eq-lca -- an equivalence, not an approximation, and the one @sec-scoring
relies on to score partial responders.

Because the pseudo-likelihood of a finite mixture can be multimodal with
near-tied optima, each candidate is fitted from many random starts and the
reported solution is the best mode found. Starting values are generated in the
main session from a deterministic seed sequence and passed to parallel workers
as data, so no worker touches the random number generator and the result is
identical under sequential and parallel execution and under any number of
workers. This matters more than it appears: a search that ranks model sizes on
inconsistently located maxima is ranking the optimiser rather than the models.

Class labels are arbitrary, so any two fits may agree completely and still not
be comparable coordinate by coordinate. We align a fit to a reference by
minimising the total squared distance between their response profiles over
permutations,

$$\hat\sigma = \arg\min_{\sigma\in\mathfrak S_K}\sum_{k=1}^{K}
\big\lVert \boldsymbol\rho^{\text{fit}}_{\cdot\cdot\sigma(k)}
- \boldsymbol\rho^{\text{ref}}_{\cdot\cdot k}\big\rVert^2 ,$$ {#eq-align}

solved exactly by the Hungarian algorithm [@kuhn1955; @hornik2005] rather than
by heuristic matching. Alignment is applied to every replicate refit before it
enters the variance, and to any comparison across fits.

One internal check is worth naming because it costs almost nothing and catches
the class of error nothing else here would. At unit weights @eq-pseudo reduces
to the ordinary log-likelihood, so the weighted implementation and an
independent unweighted one --- `poLCA` [@linzer2011] serves --- are maximising
the same function and must agree up to optimiser tolerance and mode selection.
Run as an assertion rather than an assumption, with each solution warm-started
at the other's optimum so the comparison is between optima rather than between
stopping rules, it separates the possible causes of any gap: equal
log-likelihoods with different parameters indicate drift along a flat ridge,
and unequal ones say which search found the better mode. It validates
estimation and says nothing about the variance implementation, which is a
separate concern.

# Design-based variance {#sec-variance}

All uncertainty reported here comes from stratified delete-one-PSU jackknife
replication [@wolter2007; @rust1996; @lumley2004]. With $a_h$ primary sampling
units in stratum $h$, the replicate deleting PSU $\alpha$ of stratum $h$
assigns weight zero to that unit's respondents, multiplies the weights of the
remaining respondents in stratum $h$ by $a_h/(a_h-1)$, and leaves other strata
unchanged. Writing $\hat{\boldsymbol\theta}_{(h\alpha)}$ for the estimator
recomputed on those weights,

$$\widehat{\mathbf V}(\hat{\boldsymbol\theta}) = \sum_{h=1}^{H}\frac{a_h-1}{a_h}
\sum_{\alpha=1}^{a_h}
\big(\hat{\boldsymbol\theta}_{(h\alpha)}-\hat{\boldsymbol\theta}\big)
\big(\hat{\boldsymbol\theta}_{(h\alpha)}-\hat{\boldsymbol\theta}\big)^{\!\top},
\qquad
\nu = \sum_{h} a_h - H ,$$ {#eq-jkn}

with $\nu$ the design degrees of freedom used for every $t$ reference in this
paper rather than a normal quantile.

Replication is a choice here and not a necessity: a pseudo-likelihood does
admit a sandwich variance [@binder1983], which is what Mplus and Stata compute
under survey estimation. We choose replication for three reasons specific to
this problem. @eq-jkn propagates to new statistics without rederivation, so the
domain estimates of @sec-domains and the corrections applied to them inherit
design-based variance by recomputation where linearization would require a
fresh delta-method derivation at each stage. It avoids inverting an information
matrix of several hundred dimensions with many parameters at or near the
boundary. And it makes the design specification auditable, since a misspecified
cluster identifier changes the replicate structure visibly rather than quietly.

A stratum containing a single PSU cannot take the $a_h/(a_h-1)$ scaling. The
`survey.lonely.psu` option governs linearization variance and does not change
how replicate weights are constructed, so a singleton stratum would contribute
no variance and the resulting standard errors would be anticonservative in
exactly the strata where the data are thinnest. Singleton strata are therefore
a hard stop, and the check runs on the analysis frame rather than the released
file, because case exclusions can create singletons the provider's data does
not have.

Intervals for a probability are formed on the logit scale by the delta method
rather than as $\hat p \pm t\,\mathrm{se}$. Since
$\mathrm{d}\,\mathrm{logit}(p)/\mathrm{d}p = 1/\{p(1-p)\}$,

$$\mathrm{se}_{\mathrm{logit}}(\hat p) = \frac{\mathrm{se}(\hat p)}{\hat p(1-\hat p)},
\qquad
\mathrm{CI} = \mathrm{logit}^{-1}\!\big(\mathrm{logit}(\hat p) \pm t_{0.975,\nu}\,
\mathrm{se}_{\mathrm{logit}}(\hat p)\big),$$ {#eq-logit}

which keeps the interval inside $(0,1)$ without truncation and improves
coverage near the boundary. Two cases are excluded from the transform: an
estimate within $10^{-3}$ of a boundary has no meaningful logit-scale standard
error, and a bias-corrected share can legitimately fall outside $[0,1]$, where
truncation would misstate it. Both keep a Wald interval and are flagged rather
than mixed in with the rest.

# The analysis frame {#sec-frames}

Everything that follows is computed on one frame: the respondents who answered
every item in the battery, carrying the strata, primary sampling units and base
weights the provider released.

```{r frames}
cfg_l   <- arm_lca$cfg
items_l <- cfg_l$items
cats_l  <- arm_lca$cats
dat_l   <- cfg_l$data
w_l     <- dat_l[[cfg_l$weight]]
n_l     <- nrow(dat_l)
scale_l <- n_l / sum(w_l)

d_l <- build_rep_design(dat_l, cfg_l)
des_l <- d_l$des; rep_l <- d_l$rep_des
nu_l <- degf(rep_l); crit_l <- qt(0.975, nu_l)

init_parallel(cfg_l)
```

```{r frames-table}
uwe <- function(w) 1 + (sd(w) / mean(w))^2

tibble(Items = length(items_l), n = n_l,
       `CV(w)` = round(sd(w_l) / mean(w_l), 3),
       UWE = round(uwe(w_l), 2),
       Replicates = ncol(rep_l$repweights),
       `Design df` = nu_l) |>
  lca_table(align = "rrrrrr", widths = fit_widths(6, first = 1.2),
            caption = "The analysis frame. UWE is the unequal weighting effect,
                       the multiplicative variance penalty weight variation
                       imposes even when the weights are unrelated to the
                       survey variable.")
```

The unequal weighting effect $1+\mathrm{CV}^2(w)$ [@kish1965] is reported here
rather than later because it sets expectations for everything downstream. Where
it is close to one, weighted and unweighted point estimates will differ little,
and the design's contribution to the variance will be clustering rather than
weighting. Reporting it prevents the common error of attributing a design
effect to the weights on a file where the weights are nearly constant.



# Choosing the dimension {#sec-dimension}

The number of classes is never selected automatically. It is an analyst
decision recorded in configuration, taken after reading the evidence below, and
the reason is not methodological modesty. Every automatic rule available here
--- minimum BIC, the first solution that fits --- optimises one criterion while
the decision depends on several, one of which (does the solution mean
anything) no criterion measures. What the workflow owes the analyst is the
evidence in a form that cannot be misread, and a record of what was chosen.

## Enumerating classes {#sec-enumeration}

An information criterion compares a maximised log-likelihood against a penalty
in units of $\log n$, and the two must sit on the same scale. Base weights
rarely sum to $n$. Because $\ell_w$ in @eq-pseudo is linear in the weights,
multiplying it by $n/\sum_i w_i$ is exactly equivalent to refitting with
weights rescaled to sum to $n$: the point estimates are untouched and only the
criteria change. With $\ell^*_w = \ell_w \cdot n/\sum_i w_i$,

$$\mathrm{AIC} = -2\ell^*_w + 2 d_K,
\qquad
\mathrm{BIC} = -2\ell^*_w + d_K \log n,
\qquad
d_K = (K-1) + K\sum_{j=1}^{J}(C_j-1) .$$ {#eq-ic}

Without that rescaling, weights summing to substantially more than $n$ inflate
the log-likelihood against a fixed penalty and BIC leans toward too many
classes [@lumley2015]. This is a common and silent error: nothing in the output
of a weighted fit announces which scale its log-likelihood is on.

Model dimension $d_K$ grows linearly in $K$ with slope $\sum_j(C_j-1)$, so a
battery of seven-category items becomes expensive quickly. The
`r length(items_l)` items of the class arm carry
$\sum_j(C_j-1) =$ `r sum(cats_l - 1)`, giving `r df_k(K_star, cats_l)` free
parameters at $K =$ `r K_star`. A battery of eleven seven-point items would
carry 66 and 334 free parameters at five classes, which is why the choice of
battery and the choice of model are not separable questions.

Alongside the criteria we report the relative entropy of the posterior
classification, on the weighted scale,

$$R^2_E = 1 - \frac{-\sum_i w_i \sum_k p_{ik}\log p_{ik}}
{\big(\sum_i w_i\big)\log K} ,$$ {#eq-entropy}

which indexes classification sharpness [@ramaswamy1993]. The numerator is the
weighted average posterior uncertainty and the denominator its maximum,
attained when every posterior is uniform over the $K$ classes, so $R^2_E$ runs
from zero at no classification information to one at perfect separation; values
near 0.8 and above are conventionally read as good separation. Weighting it is
not cosmetic: an unweighted entropy would describe the achieved sample's
classification sharpness while every other quantity in the analysis describes
the population.

The bootstrap likelihood ratio test, often preferred for class enumeration
under simple random sampling [@nylund2007], is omitted deliberately. Its
parametric bootstrap draws new samples of independent observations from the
fitted model, which is not the process that generated a stratified clustered
sample. Mplus disables the test under complex-design analysis for the same
reason, and nothing is gained by reporting a $p$-value whose null distribution
is known to be wrong. We are not aware of a design-based replacement for it,
and treat its absence as a genuine gap rather than one this workflow closes.

```{r enumerate}
enum <- future_map(cfg_l$K_range, function(K) {
  fit <- fit_lca(dat_l, w_l, cats_l, items_l, K, start_seeds(cfg_l, K))
  ll_n <- fit$ll * scale_l
  d <- df_k(K, cats_l)
  tibble(K = K, logLik = ll_n, df = d,
         AIC = -2 * ll_n + 2 * d,
         BIC = -2 * ll_n + d * log(n_l),
         entropy = entropy_R2(fit$post, w_l, K),
         converged = fit$converged)
}, .options = furrr_options(seed = NULL,
     packages = c("matrixStats", "clue", "purrr", "dplyr", "tidyr"))) |>
  list_rbind()

enum |>
  mutate(across(c(logLik, AIC, BIC), function(x) round(x, 1)),
         entropy = round(entropy, 3)) |>
  lca_table(align = "r",
            caption = "Weighted fit statistics by number of classes, on the
                       sum-to-n scale. Nothing is selected here.")
```

```{r fig-enum}
#| fig-height: 2.8
#| fig-cap: "Enumeration evidence. The criterion penalises model dimension; the entropy indexes classification sharpness. Neither decides alone, and they routinely point in different directions."
enum |>
  select(K, BIC, entropy) |>
  pivot_longer(-K, names_to = "criterion", values_to = "value") |>
  mutate(criterion = if_else(criterion == "BIC", "Weighted BIC",
                             "Relative entropy")) |>
  ggplot(aes(K, value)) +
  geom_line(linewidth = 0.6, colour = viridisLite::viridis(1, begin = 0.4)) +
  geom_point(size = 2.2, colour = viridisLite::viridis(1, begin = 0.4)) +
  facet_wrap(~ criterion, scales = "free_y") +
  scale_x_continuous(breaks = cfg_l$K_range) +
  labs(x = "Number of classes (K)", y = NULL) +
  theme_lca()
```

# The fitted model {#sec-fitted}

```{r fit-class}
fit_l <- fit_lca(dat_l, w_l, cats_l, items_l, K_star,
                 start_seeds(cfg_l, K_star), maxit = 5000L, tol = 1e-10)

inp_l    <- make_inputs(dat_l, items_l, cats_l)
post_l   <- posterior_of(fit_l$pi, fit_l$rho, inp_l$Y)
colnames(post_l) <- paste0("post_segment", seq_len(K_star))
modal_l  <- max.col(post_l, ties.method = "first")

glue("Class arm: K = {K_star} on {n_l} cases from {cfg_l$n_starts} starts. ",
     "Converged: {if (isTRUE(fit_l$converged)) 'yes' else 'NO'}. ",
     "Weighted log-likelihood (sum-to-n scale) {round(fit_l$ll * scale_l, 1)}. ",
     "Weighted relative entropy {round(entropy_R2(post_l, w_l, K_star), 3)}, ",
     "mean maximum posterior ",
     "{round(mean(post_l[cbind(seq_len(n_l), modal_l)]), 3)}.")
```

```{r level-pattern}
lpr <- level_pattern_ratio(fit_l, items_l)

lpr |>
  mutate(across(everything(), function(x) round(x, 3))) |>
  lca_table(align = "rrr",
            col.names = c("SD of class level", "SD of class pattern", "Ratio"),
            widths = c("14em", "14em", "8em"),
            caption = "The level-to-pattern ratio of @eq-ratio for the fitted
                       class solution. Well below one says the classes differ
                       mainly in how high they answer, which a factor model
                       would describe with far fewer parameters. Near or above
                       one says they reorder the items.")
```

For this battery the ratio is `r round(lpr$ratio, 2)`, which sits
`r if (lpr$ratio >= 1) "at or above" else "below"` one: the classes differ
`r if (lpr$ratio >= 1) "by reordering the items rather than by sitting at different heights on them" else "mainly in how high they answer overall, with less reordering of the items"`.
A mixed-format battery spanning durable goods, food insecurity, crime
victimisation and migration intention has no reason to behave like a single
continuum, and a ratio at or above one is the quantitative form of the
judgement that a class model rather than a factor model belongs on it. A ratio
well below one on such a battery is the more interesting result, because it
says the classes are approximating a continuum with a staircase and that a
factor model would describe the same structure with far fewer parameters.

The ratio is evidence and not a test. It has no null distribution and no
threshold, it is computed from a fitted solution and therefore inherits
whatever is wrong with that solution, and it is read alongside the eigenvalues
and alongside what the analyst knows about the questions.

# Fit diagnostics {#sec-diagnostics}

A model is diagnosed before it is interpreted, and the diagnostics asked here
are design-aware where a design-aware version exists and honest about ranking
rather than testing where one does not.

## Does each item earn its place

For the class arm, how much an item separates the classes is measured as the
mean over class pairs of the total variation distance between their conditional
response distributions,

$$D_j = \frac{2}{K(K-1)}\sum_{k<l}\frac{1}{2}\sum_{c=1}^{C_j}
\big|\rho_{jck} - \rho_{jcl}\big| \in [0,1] .$$ {#eq-discrim}

The inner term is the total variation distance between two discrete
distributions: the largest difference in probability the two assign to any
event. The leading factor is the reciprocal of the number of pairs, so $D_j$ is
a mean of quantities each in $[0,1]$. Zero means every class answers item $j$
identically, making it a candidate for removal in a sensitivity refit; near one
means the item alone nearly determines membership.

Two properties make @eq-discrim the right choice for a mixed-format battery.
It is bounded in $[0,1]$ whatever the number of response categories, so a
binary and a seven-point item are on one scale. And it is sensitive to shape
rather than only to location: a class that avoids the middle of a scale
registers here even when its average answer sits where everyone else's does,
which suits a model that treats the categories as unordered. Reported beside it
is the range of the expected response, scaled by the number of categories,
which is what an analyst reads off a profile plot and which does assume the
categories are ordered. The two usually agree; where they do not, the item is
doing something worth looking at.

```{r item-discrimination}
item_discrimination(fit_l, items_l) |>
  mutate(across(c(discrimination, range), function(x) round(x, 3))) |>
  lca_table(col.names = c("Item", "Discrimination", "Range"), align = "lrr",
            widths = c("14em", "9em", "6em"),
            caption = "Two views of how much each item separates the classes.
                       Where they disagree, the item is worth inspecting rather
                       than dropping.")
```

## Local independence in the class arm

Local independence is probed pairwise. For items $a$ and $b$, let
$\widehat P_{ab}(c,d)$ be the design-weighted observed proportion giving
response $c$ to $a$ and $d$ to $b$, and let

$$\Pi_{ab} = \boldsymbol\rho_a \operatorname{diag}(\boldsymbol\pi)\,
\boldsymbol\rho_b^{\top}$$ {#eq-implied}

be the model-implied joint distribution, with $\boldsymbol\rho_a$ the
$C_a\times K$ matrix of conditional probabilities for item $a$. @eq-implied is
the model's own statement about the pair: condition on class, multiply the two
conditional distributions because the model asserts independence given class,
and average over classes weighted by prevalence. The bivariate residual is the
total variation distance between observed and implied,

$$\mathrm{BVR}_{ab} = \tfrac{1}{2}\sum_{c,d}
\big|\widehat P_{ab}(c,d) - \Pi_{ab}(c,d)\big| \in [0,1] .$$ {#eq-bvr}

This departs from the Pearson-type statistic conventionally reported, and the
departure is deliberate. @oberski2013 showed that the Pearson statistic's null
distribution is already not chi-square for unweighted binary data, and
recommended bootstrap calibration; under design weighting and a
pseudo-likelihood on a clustered sample, no reference distribution is valid at
all. A Rao--Scott design adjustment would repair the design component but not
the fact that the null is not chi-square even under simple random sampling, so
it does not rescue the statistic here.

A statistic without a null distribution is an effect size, and an effect size
should sit on an interpretable scale. @eq-bvr is bounded in $[0,1]$, is exactly
zero under perfect local independence, reads as the largest probability
discrepancy over any event in the two-way table, and shares its scale with
@eq-discrim. It also removes a numerical failure of the Pearson form, which
divides by the model-implied cell probability and so can be dominated by a
single near-empty corner of a large table. What it gives up is a decision rule:
it ranks pairs and does not test them, and the analyst reads the ranking
against the questionnaire.

```{r bvr}
bvr_tbl <- bvr_pairs(dat_l, w_l, items_l, fit_l)

bvr_tbl |>
  mutate(bvr = round(bvr, 3)) |>
  slice_head(n = 8) |>
  lca_table(align = "llr", col.names = c("Item A", "Item B", "BVR"),
            widths = c("14em", "14em", "6em"),
            caption = "Eight largest bivariate residuals. A ranking, not a
                       test: no reference distribution is valid under a
                       design-weighted pseudo-likelihood.")
```

# What the design costs the measurement model {#sec-deff}

The replicate variance of @eq-jkn is applied to the fitted parameters of both
arms by refitting inside every replicate. This is the expensive step of the
workflow and it is also the one that produces the number nobody can obtain any
other way: the design effect on a measurement model parameter,

$$\mathrm{deff}(\hat\theta) =
\frac{\widehat{\mathrm{V}}_{\text{design}}(\hat\theta)}
     {\widehat{\mathrm{V}}_{\text{model}}(\hat\theta)} ,$$ {#eq-deff}

the ratio of the replicate variance to the variance the estimator reports under
independence. A deff of two means the model-based interval is $\sqrt2$ times
too narrow, and every downstream comparison built on that parameter inherits
the error.

The two arms differ in one operational respect. The class arm warm-starts each
replicate from the full-sample solution and aligns it by @eq-align before it
enters the variance, so the resulting variance is **within-mode** by
construction: it expresses design variance around the reported solution, not
uncertainty about which mode the likelihood surface should have selected. That
is a real limitation and it is stated rather than hidden. Replicates that fail
to converge are counted and disclosed rather than dropped quietly: treating a
failed replicate as a zero deviation biases the variance downward, and dropping
it without saying so understates it.

```{r jackknife-class}
param_meta <- bind_rows(
  tibble(kind = "pi", item = NA_character_, category = NA_integer_,
         segment = seq_len(K_star)),
  imap(cats_l, function(Cj, item_name)
    expand_grid(segment = seq_len(K_star), category = seq_len(Cj)) |>
      transmute(kind = "rho", item = item_name, category, segment)) |>
    list_rbind())

flatten_fit <- function(fit) c(fit$pi, unlist(map(fit$rho, as.vector)))

theta_lca <- function(w_rep) {
  fit <- em_run(inp_l$Y, inp_l$OH, cats_l, w_rep, K_star,
                init = fit_l[c("pi", "rho")], maxit = 1000L, tol = 1e-10)
  flatten_fit(align_to(fit, fit_l))
}

est_l <- flatten_fit(fit_l)
se_l  <- sqrt(diag(replicate_variance(rep_l, theta_lca, est_l)))

ci_l <- param_meta |>
  mutate(estimate = as.numeric(est_l), se = as.numeric(se_l),
         boundary = estimate < 1e-3 | estimate > 1 - 1e-3,
         p_c = pmin(pmax(estimate, 1e-6), 1 - 1e-6),
         gse = se / (p_c * (1 - p_c)),
         lo = if_else(boundary, pmax(0, estimate - crit_l * se),
                      plogis(qlogis(p_c) - crit_l * gse)),
         hi = if_else(boundary, pmin(1, estimate + crit_l * se),
                      plogis(qlogis(p_c) + crit_l * gse))) |>
  select(-p_c, -gse)

ci_l |>
  filter(kind == "pi") |>
  transmute(Class = segment,
            Prevalence = sprintf("%.3f", estimate),
            `95% CI` = sprintf("[%.3f, %.3f]", lo, hi),
            SE = sprintf("%.4f", se)) |>
  lca_table(align = "lrrr", widths = c("6em", "10em", "14em", "8em"),
            caption = "Class prevalences with design-based intervals on the
                       logit scale.")
```

The reading of these tables is the point of computing them, and it has to be
done on each file rather than carried over. Where the unequal weighting effect
of @sec-frames is close to one, weighting moves almost nothing and the design's
whole contribution is clustering; where the median design effect on the
loadings then also sits near one, a model-based analysis of that battery on
that file would have been approximately right. That is a finding, demonstrated,
and not an entitlement to assume the same elsewhere. On a file with informative
weights, or on a battery whose answers concentrate within clusters, the same
table is what tells the analyst that the standard errors they were about to
publish were a fraction of what they should have been. Nothing short of
computing it distinguishes the two cases, which is the argument for computing
it every time.

# Scoring {#sec-scoring}

Both arms turn a fitted measurement model into a per-respondent quantity that
the domain estimates of @sec-domains average. The two scoring steps have
different failure modes, and both failure modes are attenuation or its
correction.

## Posterior membership, and partial responders

Local independence makes assignment exact rather than approximate for
respondents who skipped items. A missing answer drops out of the within-class
product, so for respondent $i$ with answered item set $A_i$,

$$p^{\mathrm{obs}}_{ik} = \frac{\pi_k \prod_{j\in A_i}\rho_{j\,y_{ij}\,k}}
{\sum_{l}\pi_l \prod_{j\in A_i}\rho_{j\,y_{ij}\,l}},
\qquad
\hat X_i = \arg\max_k p^{\mathrm{obs}}_{ik} .$$ {#eq-score}

@eq-score is the exact posterior given what respondent $i$ actually answered,
under the model and under ignorability of the missingness pattern given class.
An information floor is enforced: respondents answering fewer than
`r cfg_l$min_items` items receive no assignment rather than a weakly informed
one.

Scoring precedes domain estimation and all domain estimation runs on the scored
frame rather than on the estimation sample. This follows the logic of
three-step latent class analysis, and it matters substantively rather than
cosmetically: item nonresponse on an attitude battery correlates with
education, age and other characteristics of interest, so restricting domain
estimation to complete responders introduces a selection into exactly the
composition being estimated. The cost is that respondents scored from fewer
items carry flatter posteriors and therefore more classification error, which
is reported rather than assumed away. The replicate design is rebuilt on the
scored frame and the singleton-stratum check is repeated there, because the
frame has changed.

```{r score-class}
pred <- predict_segments(arm_lca$survey_dat_full, fit_l, items_l,
                         cfg_l$min_items)
survey_scored <- bind_cols(arm_lca$survey_dat_full, pred)
post_cols <- paste0("post_segment", seq_len(K_star))

# A demographic level whose last respondent went unscored has to go with them:
#   svyby returns no row for an empty level while count(.drop = FALSE) returns
#   one, and the three estimator frames of @sec-domains would stop lining up.
scored_l <- survey_scored |>
  filter(!is.na(segment)) |>
  mutate(segment = factor(segment, levels = seq_len(K_star))) |>
  drop_empty_levels(cfg_l$aux, "scored frame")

ds_l <- build_rep_design(scored_l, cfg_l)
des_sl <- ds_l$des; rep_sl <- ds_l$rep_des
nu_sl <- degf(rep_sl); crit_sl <- qt(0.975, nu_sl)

glue("Estimation sample {n_l} item-complete cases; scored sample ",
     "{nrow(scored_l)} of {nrow(survey_scored)} respondents at a floor of ",
     "{cfg_l$min_items} items, a gain of {nrow(scored_l) - n_l}. ",
     "Domain design: {ncol(rep_sl$repweights)} replicates, {nu_sl} degrees of ",
     "freedom.")
```

```{r score-quality}
scored_l |>
  group_by(n_items_answered) |>
  summarise(n = n(), mean_max_posterior = round(mean(max_posterior), 3),
            .groups = "drop") |>
  lca_table(align = "rrr", widths = c("12em", "6em", "14em"),
            col.names = c("Items answered", "n", "Mean max posterior"),
            caption = "Assignment quality by number of items answered.
                       Respondents scored from fewer items carry flatter
                       posteriors and so more classification error; a sharp
                       fall at the low end argues for raising the floor.")
```

# Domain estimation {#sec-domains}

The quantity a substantive reader wants is not the measurement model. It is
what the latent variable looks like within demographic domains: which kinds of
respondent are concentrated where, or which groups sit higher on the factor.
Getting there from a fitted measurement model is where most of the error in
practice is introduced, for a reason that is easy to state and easy to forget.
**The latent variable is estimated, not observed.** Cross-tabulating an
estimate of it against a demographic inherits the estimate's error, and that
error does not average out; it attenuates.

## Three estimators, reported together

Each domain table carries the same quantity three ways, so the two corrections
can be read separately rather than compounded into a single number nobody can
decompose.

**Unweighted** is a plain cross-tab or group mean: sample proportions with
binomial standard errors, or group means with $s/\sqrt{n}$, as if the
respondents had been drawn independently from the whole population. This is
what a default analysis produces.

**Design-based** applies the sampling weights and takes its variance from the
replicates of @eq-jkn. For the class arm,

$$\hat p_{dk} = \frac{\sum_{i\in d} w_i \,\mathbf 1(\hat X_i = k)}
{\sum_{i\in d} w_i} ,$$ {#eq-domain}

the design-weighted share of domain $d$ assigned to class $k$. Rows sum to one across
classes by construction, which means the cells within a row are not
independent, which in turn is why the comparisons of @sec-separations use the
full replicate covariance rather than treating cells as separate estimates.

**Corrected** removes the attenuation the classification step introduces. That
is the BCH correction of @sec-bch.

The gap from the first row to the second is what ignoring the design costs. The
gap from the second to the third is what measurement error in the latent
variable costs. Reporting one number instead of three hides which of the two an
analyst is looking at.

## The BCH correction {#sec-bch}

Modal assignment is an error-prone measurement of true class, and
cross-tabulating it against anything pulls the association toward the marginal
distribution [@bolck2004; @vermunt2010]. The direction is what makes this
tractable: a difference that survives modal assignment is real, while a null
under modal assignment may be a genuine difference that classification error
erased.

Let $D$ hold the design-weighted classification error rates,
$D_{ks} = P(\hat X = s \mid X = k)$, estimated from the posteriors and the
modal assignments,

$$D_{ks} = \frac{\sum_i w_i\, p_{ik}\,\mathbf 1(\hat X_i = s)}
{\sum_{s'}\sum_i w_i\, p_{ik}\,\mathbf 1(\hat X_i = s')} .$$ {#eq-bchD}

Each respondent's hard assignment is then replaced by row $\hat X_i$ of
$D^{-1}$, and the domain estimate of @eq-domain is computed on those corrected
weights [@bakk2013; @bakk2014]. Entries of $D^{-1}$ can be negative, which is a
property of the correction rather than a fault, and rows still sum to one
because $D$'s rows do. A corrected share can therefore legitimately fall
outside $[0,1]$, which is one of the two cases @eq-logit excludes from the
logit transform: truncating it would misstate the correction rather than
improve the interval.

One implementation choice here is load-bearing and is the kind of thing that
distinguishes a correction that carries its uncertainty from one that pretends
to. The posteriors and the modal assignments are held at their full-sample
values, but $D$ and the correction weights are **rebuilt inside every
replicate**, so the intervals carry the design variance of the classification
table and of the cross-tabulation together rather than treating $D$ as known.
Treating $D$ as fixed is the common shortcut and it understates the variance of
the corrected estimate by exactly the amount that the classification table is
itself uncertain.

A singular $D$ is a hard stop rather than a pseudo-inverse. It arises when a
class takes no modal assignments inside a replicate, which is a statement that
the class is too small to survive deleting one primary sampling unit --- a
finding about the model, and one that should reach the analyst as an error
rather than be smoothed over by a generalised inverse.

```{r domains-class}
marg_l <- map(cfg_l$aux, function(v) {
  m <- svymean(reformulate(v), rep_sl, na.rm = TRUE)
  raw <- scored_l |>
    filter(!is.na(.data[[v]])) |>
    count(.data[[v]], name = "n") |>
    set_names(c("level", "n"))
  tibble(variable = v,
         level = str_remove(names(coef(m)), paste0("^", v)),
         weighted = as.numeric(coef(m)), se = as.numeric(SE(m))) |>
    left_join(mutate(raw, level = as.character(level)), by = "level")
}) |>
  list_rbind()

dom_meta <- map(cfg_l$aux, function(v)
  expand_grid(level = levels(scored_l[[v]]), segment = seq_len(K_star)) |>
    mutate(variable = v)) |>
  list_rbind() |>
  mutate(idx = row_number())

indicator_of <- function(v)
  local({
    M = outer(as.character(scored_l[[v]]), levels(scored_l[[v]]), "==") + 0
    M[is.na(M)] = 0
    M
  })
Mlist <- set_names(map(cfg_l$aux, indicator_of), cfg_l$aux)

P_fixed     <- as.matrix(scored_l[, post_cols, drop = FALSE])
modal_fixed <- as.integer(scored_l$segment)
S_fixed     <- outer(modal_fixed, seq_len(K_star), `==`) + 0

# Design-based and BCH as functions of a weight vector, so the full replicate
#   covariance between domain cells exists and @sec-separations can test
#   differences on it rather than on an interval-overlap rule.
theta_dsg <- function(w_rep)
  unlist(map(cfg_l$aux, function(v)
    as.vector(t(crossprod(Mlist[[v]], w_rep * S_fixed) /
                as.numeric(crossprod(Mlist[[v]], w_rep))))), use.names = FALSE)

theta_bch <- function(w_rep) {
  wU = w_rep * bch_weights(P_fixed, modal_fixed, w_rep)
  unlist(map(cfg_l$aux, function(v)
    as.vector(t(crossprod(Mlist[[v]], wU) /
                as.numeric(crossprod(Mlist[[v]], w_rep))))), use.names = FALSE)
}

w_samp   <- weights(des_sl, "sampling")
est_dsg  <- theta_dsg(w_samp)
V_dsg    <- replicate_variance(rep_sl, theta_dsg, est_dsg)
est_bch  <- theta_bch(w_samp)
se_bch   <- sqrt(diag(replicate_variance(rep_sl, theta_bch, est_bch)))
stopifnot(length(est_dsg) == nrow(dom_meta))

dom_dsg <- dom_meta |>
  mutate(estimator = "Design-based", p = est_dsg,
         se = sqrt(diag(V_dsg))) |>
  mutate(lo = pmax(0, p - crit_sl * se), hi = pmin(1, p + crit_sl * se))

dom_bch <- dom_meta |>
  mutate(estimator = "BCH", p = est_bch, se = se_bch,
         lo = p - crit_sl * se, hi = p + crit_sl * se)

dom_naive <- map(cfg_l$aux, function(v) {
  scored_l |>
    filter(!is.na(.data[[v]])) |>
    count(.data[[v]], segment, name = "n", .drop = FALSE) |>
    set_names(c("level", "segment", "n")) |>
    group_by(level) |>
    mutate(n_level = sum(n), p = n / n_level) |>
    ungroup() |>
    transmute(variable = v, level = as.character(level),
              segment = as.integer(segment), estimator = "Unweighted", p,
              se = sqrt(p * (1 - p) / n_level),
              lo = pmax(0, p - qnorm(0.975) * se),
              hi = pmin(1, p + qnorm(0.975) * se))
}) |>
  list_rbind()

est_levels <- c("Unweighted", "Design-based", "BCH")
domain_l <- bind_rows(dom_naive, select(dom_dsg, -idx),
                      select(dom_bch, -idx)) |>
  mutate(estimator = factor(estimator, levels = est_levels))

# The frames are joined by key rather than differenced by position: svyby
#   returns nothing for a demographic level with no cases while
#   count(.drop = FALSE) returns a row for it, so a positional difference can
#   pair the wrong cells rather than fail.
gaps <- dom_naive |>
  select(variable, level, segment, p_naive = p, se_naive = se) |>
  inner_join(select(dom_dsg, variable, level, segment, p_dsg = p, se_dsg = se),
             by = c("variable", "level", "segment")) |>
  inner_join(select(dom_bch, variable, level, segment, p_bch = p),
             by = c("variable", "level", "segment"))
stopifnot(nrow(gaps) == nrow(dom_meta))

# Cells at an observed 0 or 1 have a zero binomial standard error, so the ratio
#   there is 0/0. They say nothing about the clustering cost and are excluded
#   from the median rather than allowed to poison it.
r_se <- gaps$se_dsg / gaps$se_naive

glue("Weighting moves the class shares by a median of ",
     "{sprintf('%.4f', median(abs(gaps$p_dsg - gaps$p_naive)))}; the BCH ",
     "correction moves them a further ",
     "{sprintf('%.4f', median(abs(gaps$p_bch - gaps$p_dsg)))}, at most ",
     "{sprintf('%.4f', max(abs(gaps$p_bch - gaps$p_dsg)))}, against a median ",
     "design-based standard error of {sprintf('%.4f', median(gaps$se_dsg))}. ",
     "Design-based standard errors are a median of ",
     "{sprintf('%.2f', median(r_se[is.finite(r_se)]))} times the naive ones.")
```

## Which differences are real {#sec-separations}

An interval-overlap rule is the common way to decide whether two domain levels
differ, and it is wrong in a specific direction: two 95 percent intervals that
fail to overlap correspond to a test at an $\alpha$ well below 0.05, so the
rule is conservative and its actual level is unstated. It is also wrong in kind
here, because the cells of a domain row are estimated from the same replicates
and are correlated, and an overlap rule ignores that covariance entirely.

The replicate design already carries what is needed. For two levels $a$ and $b$
of the same demographic, on the same class or factor,

$$t = \frac{\hat\theta_a - \hat\theta_b}
{\sqrt{\widehat{\mathrm{V}}_{aa} + \widehat{\mathrm{V}}_{bb}
- 2\widehat{\mathrm{V}}_{ab}}} ,
\qquad t \sim t_\nu ,$$ {#eq-wald}

with $\widehat{\mathrm{V}}$ the replicate covariance of the domain estimates
and $\nu$ the design degrees of freedom of @eq-jkn. Within each demographic the
resulting $p$-values are Holm-adjusted across the pairs, which controls the
family-wise error rate over the comparisons actually made rather than over an
unstated number.

The tests run on the **design-based** estimator deliberately, not on the
corrected one. Modal assignment attenuates differences, so a pair that
separates under the design-based estimator separates despite the classification
error, and the corrected column can only widen the gap. This makes the
resulting claims conservative in a direction that is stated rather than
accidental.

```{r separations}
wald_l <- list(meta = dom_meta, est = est_dsg, V = V_dsg, df = nu_sl)

cat(domain_separations(domain_l, variable = "education", wald = wald_l), "\n")
```

## What the tables say

The full workflow produces a table of this form for every demographic in the
configuration. One is shown here; the rest are mechanical repetitions of it and
belong in an appendix or an application rather than in a methods paper.

```{r domain-table-class}
domain_l |>
  filter(variable == "education") |>
  transmute(Class = segment, Estimator = as.character(estimator), Level = level,
            cell = sprintf("%.3f [%.3f, %.3f]", p, lo, hi)) |>
  pivot_wider(names_from = Level, values_from = cell) |>
  arrange(Class, match(Estimator, est_levels)) |>
  lca_table(align = "llrrr", widths = fit_widths(5, first = 1.2), font_size = 7,
            caption = "Class shares within levels of education, three ways.
                       The first gap is what ignoring the design costs; the
                       second is what classification error costs.")
```

Two limits belong with any such table and are not repaired by anything above.
The latent variable is estimated rather than observed, so a difference that
appears is real while a null is not evidence of absence --- the asymmetry runs
one way and should be stated that way in the text a reader sees. And a level
resting on very few respondents produces a wide interval that is not a finding;
sample sizes belong beside every estimate for that reason.

Worth adding, because the rule is easily mistaken for a preference for caution:
a design-based interval is not always wider than a naive one. Stratification
can make it narrower, and cells above show exactly that. The rule is not
*prefer the wider interval*. It is *prefer the interval that matches how the
sample was actually drawn*.

# What this workflow does not fix {#sec-limitations}

Stated as failures rather than as caveats, because a limitation an analyst can
act on is worth more than one that reassures.

**Scores instead of a structural model.** Both arms estimate the measurement
model, extract a per-respondent quantity, and then estimate domain differences
on that quantity. This is a three-step approach, and the corrections above ---
BCH in one arm, unshrunk scores in the other --- repair its attenuation but not
its architecture. The alternative is to put the demographics inside the model
as covariates on the latent variable, where the group differences are estimated
jointly with the measurement parameters and no scoring step intervenes. That is
a different specification, and it is the right one when the domain differences
are the estimand rather than a description. The group differences are then estimated
jointly with the measurement parameters, and no scoring step intervenes.

**Within-mode variance in the class arm.** Replicate refits are warm-started
from the full-sample solution and aligned to it, so the reported variance
expresses design variance around the reported mode. It does not express
uncertainty about which mode the surface should have selected. Where the
enumeration evidence is close between adjacent $K$, or where the log-likelihood
surface has near-tied optima, that uncertainty is real and is not in any
interval reported here.

**A structure read off the same data it is tested on.** The factor structure of
@sec-fitted was chosen from an exploratory pass on this file and then fitted
confirmatorily on the same file. Its fit indices are not an independent test of
it. The honest remedy is a second sample; a resampling of this one does not
substitute.

**No design-based enumeration test.** The bootstrap likelihood ratio test is
excluded for good reason and nothing replaces it. Class enumeration here rests
on information criteria, entropy and judgement, and information criteria under
a pseudo-likelihood do not have the asymptotic justification they have under a
likelihood.

**Modification indices are model-based.** The factor arm's local independence
check ignores clustering, so it ranks candidate residual covariances under an
understated variance. A design-based version of it would be a genuine addition
and does not exist here.

**Nonresponse and coverage are untouched.** Everything above is conditional on
the released weights being the right adjustment for nonresponse and the frame
covering the target population. The workflow carries sampling variance
faithfully and says nothing about the components of total survey error that
precede it. A recode that misgroups a demographic is likewise invisible to
every diagnostic in this paper, which is why the audit of @sec-data is a table
to be read rather than a check that passes.

**One deletion is substantive and is worth naming.** In the demonstration file
the education recode maps the "no formal schooling" response to missing, which
removes the least-educated respondents from every education domain estimate
rather than reporting them as a level. Their answer was given; this is a
deletion, not missing data, and it is the analyst's call rather than a defect
--- but it is exactly the kind of decision that the recode audit exists to make
visible.

# Recommendations for practice {#sec-recommendations}

Six of the choices above generalise beyond this pair of models, and are worth
stating without their surrounding machinery.

1. **Put the log-likelihood and the penalty on the same scale before comparing
   information criteria under weights.** Rescaling the weighted log-likelihood
   by $n/\sum_i w_i$ leaves the point estimates untouched and is the difference
   between a criterion that penalises dimension and one that does not.

2. **Report the three estimators rather than the corrected one.** An analyst
   who sees only the corrected number cannot tell whether the design or the
   measurement error was doing the work, and the two have different remedies.

3. **Rebuild the correction inside every replicate.** A classification table,
   a shrinkage factor or a calibration constant treated as known understates
   the variance of the corrected estimate by exactly its own uncertainty.

4. **Compute the check rather than asserting it.** A comment in an earlier
   version of this workflow described a sensitivity check as having been run.
   It had not been. When it was, four domain cells turned out to be carrying a
   handful of respondents rather than describing their groups. Every assertion
   of the form *this is small enough not to matter* is a computation somebody
   has decided not to run.

5. **Make silent failures loud.** A singleton stratum, a singular
   classification table, a demographic level that exists in one frame and not
   in another, a sampling weight in a loading table --- each of these produces
   a plausible number rather than an error. Every one of them is a hard stop in
   this workflow, and each stop was added after the corresponding silent
   failure had already happened once.

6. **Record the dimension decision in configuration, not in code.** Neither $K$
   nor the number of factors is selected automatically here. The evidence is
   produced, the analyst reads it, and the decision is written down where it
   can be audited and where a re-render cannot quietly change it.

# Reproducibility {#sec-reproducibility}

Every quantity in this paper is computed at render time from the configuration
files and the estimation code, and nothing is transcribed. Random starts are
generated in the main session from a deterministic seed sequence and passed to
parallel workers as data, so no worker touches the random number generator and
results are identical under sequential and parallel execution and under any
number of workers. Verified rather than asserted: the fitted
solution is bit-for-bit identical at one worker and at four.

```{r session}
tibble(Component = c("R", "lavaan", "survey", "Design", "Replicates",
                     "Design df (class / factor)"),
       Value = c(paste(R.version$major, R.version$minor, sep = "."),
                 as.character(packageVersion("lavaan")),
                 as.character(packageVersion("survey")),
                 "Stratified, delete-one-PSU jackknife (JKn)",
                 paste(ncol(rep_l$repweights), "/", ncol(rep_f$repweights)),
                 paste(nu_l, "/", nu_f))) |>
  lca_table(align = "ll", widths = c("18em", "24em"),
            caption = "Software and design used for every estimate in this
                       paper.")
```

# References {.unnumbered}
